A Linear Solution of the Linearity Challenge

### 2025-12-27 paper/related.tex
revised related

\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from this work
to formalise a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure depends
on the content of messages.

@@

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalisation of interpreters for models
supported by linear type systems. One of the presented case studies is the
formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. The
separating conjunction introduced in \Cref{sec:context-agda} and the typing of
the constructors for the representation of processes in \Cref{sec:process-agda}
have been heavily inspired by this work.

@@

All the formalisations mentioned so far make use of context splitting. In
contrast, \citet{ZalakainDardha21} formalise a generalisation of the linear
$\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
channel sharing/linearity) and that is based on \emph{leftover
typing}~\cite{Allais17}. 
%
With leftover typing the judgments have the form $\wtl{P}\ContextC\ContextD$
where a process $P$ is typed with respect to an input context $\ContextC$, which
describes the overall available channels, and a context of leftovers
$\ContextD$, which describes the residual channels not consumed by the process.
In this way, it is possible to ``concatenate'' typing judgments by matching the
leftovers in one judgment with the input context of the subsequent one, with no
need for splitting. As \citet{ZalakainDardha21} nicely summarise, ``context
splits contain usage information that is already present in processes,'' and
therefore can be omitted. This is true provided that channels are named
(\citet{ZalakainDardha21} use de Bruijn indices to this aim). In fact, the co-de
Bruijn representation of processes~\cite{McBride18}, whereby channels are
nameless and context splitting is performed eagerly, can be seen as the ``dual
approach'' of leftover typing: channel names provide information that is already
present in their (singleton) typing context, hence they can be omitted from
contexts and processes.

@@

    \begin{tabular}{@{}r@{\T}c@{\T}c@{\T}c@{\T}|@{\T}c@{\T}c@{\T}c@{\T}c@{\T}|@{\T}r@{\T}r@{\T}r@{\T}r@{\T}r@{\,}l@{}}
        \textbf{Formalisation} &
        \textlabel{Proof assistant} &

@@

        \textlabel{Cut} &
        \textlabel{Channel linearity} &
        \textlabel{Channel names} &

@@

        this work &

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations and
also our own. The four leftmost columns list the reference papers (sorted by
publication date), the used proof assistant, the formalised calculus, and
whether the calculus features a cut in the style of linear logic, as opposed to
separate restriction and parallel composition.
%
For simplicity, we have grouped the calculi into classes identified by an
acronym. The acronyms \CP~\cite{Wadler14} and
\GV~\cite{GayVasconcelos10,Wadler14} refer to well-known calculi in the
literature on session types. The acronym \SessionPI refers to (variants of) the
session-based $\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while
\LinearPI refers to (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the structural
version of \CP introduced by \citet{SanoKavanaghPientka23} and \Calculus is our
calculus.
%
It should be noted that the actual calculus being formalised usually differs
from (typically, is a strict subset of) the one identified by the acronym and
that the same acronym may identify different calculi. For example, the first
version of \GV described by \citet{GayVasconcelos10} is not based on linear
logic, whereas later versions~\cite{Wadler14} used \eg by
\citet{JacobsBalzerKrebbers22} are.
%
In any case, the acronym is meant to give an idea of the overall flavour of the
calculus.

The four middle columns of \Cref{tab:comparison} report the relevant qualitative
aspects of the formalisations, namely the technique for the management of typing
contexts, the representation of channels, whether the formalisation is
intrinsically or extrinsically typed, and whether the calculus makes use of
continuation channels. Concerning the use of continuations, the approaches based
directly on the linear $\pi$-calculus (into which sessions can be encoded) are
marked with {\YES} and those based on a calculus with native sessions are marked
with {\NO}. The calculus \SCP is marked with {\MAYBE} because, while not
directly inspired to the linear $\pi$-calculus, it makes use of explicit
continuations for defining the predicates that check the linear usage of
channels. Finally, all the approaches based on \GV are also marked with \MAYBE.
In principle, \emph{none} of these calculi makes use of continuations. However,
\GV is designed in such a way that each operation acting on a channel $s$ is a
function that returns the result of the operation (if any) \emph{along with the
same channel $s$}. In this way, the type of $s$ can be conveniently ``updated''
to take into account the effect of the operation. As observed by
\citet{Padovani17}, this semantics of the communication primitives is virtually
indistinguishable from one making use of explicit continuation channels.

The rightmost columns of \Cref{tab:comparison} report the size (in kilobytes) of
the formalisations as rough (and possibly questionable) estimates of their
complexity. Instead of counting line numbers, which may be affected by code
indentation styles and syntactical constraints of the proof assistant being
used, we have counted the number of characters after comments have been removed
and spaces have been squeezed.\footnote{Sequences of two or more consecutive
space-like characters are collapsed into a single space. The squeezing is
obtained by running the command \texttt{tr -s [:space:] file} on Unix-like
systems.}
%
The reported sizes account for the source code of the formalisation excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24}.
%
Some
formalisations~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,ZackonEtAl25}
define \emph{libraries} which can be reused in different contexts. In these
cases, the size of the library is reported separately next to the size of the
part of the development that uses it.

@@

Based on the available data, we can make the following observations.
%
The adoption of context splitting, which is very well represented, does not seem
to be a good indicator of the complexity of the formalisation. Indeed, the
formalisations based on context splitting span the whole range of sizes, from
the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
deadlock freedom) which is also the only one supporting all the features of \CP.

@@

We found no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice helps reducing the overall number of
datatypes to be defined and makes some results trivial (\eg
\Cref{thm:type-safety} formalised by \AgdaFunction{type-safety}), but the
definitions are also more involved because they incorporate invariants and the
proofs of typing preservation. It is plausible to speculate that the effort for
representing processes, types and typing rules is not substantially different
overall.

@@

Finally, it seems that the use of (explicit) continuations affects the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, no matter if they
are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
ones are 263kb on average. This observation is in agreement with the
considerations we made earlier in this paper concerning the difference between
\emph{updating} bindings versus \emph{removing and adding} bindings in a typing
context. The use of continuations enables a cleaner management of typing
contexts since linear channels become true ``use-once'' resources and there is
no need for strong updates of their type.

% It is also interesting to note that whereby \SCP, which the
% authors~\citet{SanoKavanaghPientka23} do not relate directly to the linear
% $\pi$-calculus, relies on explicit continuations to tame the complexity of
% resource management. Our comparison suggests that this aspect, more than the
% structural treatment of contexts,

% \citet{SanoKavanaghPientka23} do not relate it directly to the linear
% $\pi$-calculus.

% più che il trattamento strutturale dei contesti (ed evitare lo splitting) è
% l'uso delle continuazione a fare la differenza

>>>>

\citet{Thiemann19} formalises a subset of \GV~\citep{GayVasconcelos10}, a
functional language extended with session communication primitives, along with
an interpreter. \citet{CicconePadovani20} have taken inspiration from this work
to formalise a variant of the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99} that supports dependent types, so
as to enable the description of communication protocols whose structure may
depend on the content of messages.

@@

\citet{RouvoetPoulsenKrebbersVisser20} present a library of abstractions
inspired to separation logic aiding the formalisation of interpreters for
languages with linear resources. One of the presented case studies is the
formalisation of a fragment of \GV~\citep{GayVasconcelos10,Wadler14}. The
separating conjunction introduced in \Cref{sec:context-agda} and the typing of
the constructors for the representation of processes in \Cref{sec:process-agda}
have been adapted from this work.

@@

All the formalisations mentioned so far make use of context splitting. In
contrast, \citet{ZalakainDardha21} formalise a generalisation of the linear
$\pi$-calculus which is parametric in a \emph{usage algebra} (to account for
channel sharing/linearity) and that is based on \emph{leftover
typing}~\cite{Allais17}. 
%
Typing judgments have the form $\wtl{P}\ContextC\ContextD$ so that a process $P$
is typed with respect to an input context $\ContextC$, which describes all the
available channels, and a context of leftovers $\ContextD$, which describes the
residual channels not consumed by the process. In this way, it is possible to
``concatenate'' typing judgments by matching the leftovers in one judgment with
the input context of the subsequent one, with no need for splitting. As
\citet{ZalakainDardha21} nicely summarise, context splittings are not necessary
because they ``contain usage information that is already present in processes.''
This is true provided that channels are named (\citet{ZalakainDardha21} use de
Bruijn indices to this aim). In fact, the co-de Bruijn representation of
processes~\cite{McBride18}, whereby channels are nameless and context splitting
is performed eagerly, can be seen as the ``dual approach'' of leftover typing:
channel names provide information that is already present in their (singleton)
typing context, hence they can be omitted from contexts and processes.

@@

    \begin{tabular}{@{}l@{\T}c@{\T}c@{\T}c@{\T}c@{\T}|c@{\T}c@{\T}c@{\T}|@{\T}r@{\T}r@{\T}r@{\T}r@{\T}r@{\,}l@{}}
        \text{Reference paper} &
        \text{Prover} &

@@

        \textlabel{Cuts} &
        \text{Linearity} &
        \text{Channels} &

@@

        this &

@@

\Cref{tab:comparison} reports a summary of the aforementioned formalisations
(sorted by publication date) and also our own. The first five columns identify
the calculus being formalised specifying its reference paper, the used proof
assistant and an acronym that gives an idea of the overall flavour of the
calculus.
%
The acronyms \CP~\cite{Wadler14,GayVasconcelos25} and
\GV~\cite{GayVasconcelos10,Wadler14} are well-known in the literature on session
types. The acronym \SessionPI refers to (variants of) the session-based
$\pi$-calculus presented by \citet{HondaVasconcelosKubo98} while \LinearPI
refers to (variants of) the linear
$\pi$-calculus~\cite{KobayashiPierceTurner99}. Finally, \SCP is the structural
version of \CP introduced by \citet{SanoKavanaghPientka23} and \Calculus is our
calculus.
%
We emphasize that the actual calculus being formalised usually differs from
(typically, is a strict subset of) the one identified by the acronym and that
the same acronym may refer to different versions of the same calculus.
%
We also indicate whether the calculus features \emph{cuts} (that is, the
combination of restriction and parallel composition in correspondence with the
cut of linear logic) and continuations.

Concerning the use of continuations, the approaches based directly on the linear
$\pi$-calculus (into which sessions can be encoded) are marked with {\YES} and
those based on a calculus with native sessions are marked with {\NO}. The
calculus \SCP is marked with {\MAYBE} because, while not directly inspired to
the linear $\pi$-calculus, it makes use of explicit continuations for defining
the predicates that check the linear usage of channels. Finally, all the
approaches based on \GV are also marked with \MAYBE. In principle, \emph{none}
of these calculi makes use of continuations. However, \GV is designed in such a
way that each operation acting on a channel $s$ is a function that returns the
result of the operation (if any) \emph{along with the same channel $s$}. In this
way, the type of $s$ can be conveniently ``updated'' to take into account the
effect of the operation. As observed by \citet{Padovani17}, this semantics of
the communication primitives is virtually indistinguishable from one making use
of explicit continuation channels.

The three middle columns of \Cref{tab:comparison} report the relevant
qualitative aspects of the formalisations, namely the management of typing
contexts, the representation of channels and whether the formalisation is
intrinsically or extrinsically typed.

The rightmost columns report the size (in kilobytes) of the formalisations as
rough (and possibly questionable) estimates of their complexity. Papers
describing formalisations typically use ``lines of code'' as a measure of
development effort, but the number of lines may be affected by code indentation
styles and syntactical constraints of the proof assistant being used. For this
reason, we have preferred to count the total number of characters after comments
have been removed and spaces have been squeezed.\footnote{Sequences of two or
more consecutive space-like characters are collapsed into a single space. The
squeezing is obtained by running the command \texttt{tr -s [:space:] file} on
Unix-like systems.}
%
The reported sizes account for the source code of the formalisations excluding
examples and any safety result, if present. We have noticed that the
interpretation of ``safety'' varies widely among the formalisations and, except
for our own, they all differ from the one stated in the linearity
challenge~\citep{CarboneEtAl24}.
%
Some
formalisations~\cite{RouvoetPoulsenKrebbersVisser20,ZalakainDardha21,JacobsBalzerKrebbers22,ZackonEtAl25}
define \emph{libraries} which can be reused in different contexts. In these
cases, the size of the library is reported separately next to the size of the
part of the development that uses it.

@@

Looking on the available data, we can make the following observations.
%
The adoption of context splitting, which is very well represented, does not seem
to be a good indicator of the complexity of the formalisation. Indeed, the
formalisations based on context splitting span the whole range of sizes, from
the largest by \citet{GotoEtAl16} (543kb) to our own (21kb, without the proof of
deadlock freedom) which is also the only one supporting all the features of \CP.

@@

There is no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice helps reducing the overall number of
datatypes to be defined and makes some results trivial (\eg
\Cref{thm:type-safety} formalised by \AgdaFunction{type-safety}), but the
definitions are also more involved because they incorporate invariants and the
proofs of typing preservation. We speculate that the effort for representing
processes, types and typing rules is not substantially impacted overall.

@@

Finally, it appears that the use of (explicit) continuations is related to the
complexity of the formalisation more than anything else. Indeed, the six
smallest formalisations (excluding the deadlock freedom results) -- with an
average size of around 62kb -- are all based on continuations, no matter if they
are explicit (\LinearPI, \SCP, \LCC) or ``virtual'' (\GV), while the remaining
ones are 263kb on average. This observation agrees with the considerations made
earlier in this paper concerning the difference between \emph{updating} bindings
versus \emph{removing and adding} bindings in a typing context. The use of
continuations enables a cleaner management of typing contexts since linear
channels are true ``use-once'' resources and there is no need to update their
type.
